ChemKED: a human- and machine-readable data standard for chemical kinetics experiments

### 2017-02-26 pyked.tex
Bryan updates/notes in intro

Interestingly, this contrasts the situation for chemical kinetic models, where the
CHEMKIN~\autocite{Kee:1996ck} format dominates. Competing standards such as the
Cantera~\autocite{Cantera:2.3.0} CTI format or FlameMaster~\autocite{FlameMaster:ref,FlameMaster:code}
lag behind considerably, although use of the former continually grows due to Cantera's
open availability.
% Similarly, % with more space, discuss JANAF/NASA polynomial standard ?

Instead, such data must typically be found in non- or under-documented, nonstandard
comma-separated value files and Excel spreadsheets, or contained in PDF tables
rather than as supplementary material associated with a paper.
(In the worst cases, data is only available in figures---and in some cases not at all.)
\todo{call out specific examples of these...?}{}
These practices limit the wide use of the valuable, extensive experimental data
available in the literature. Frenklach~\autocite{Frenklach:2007bm} further explained the
benefits of standardized, widely available data for combustion.

While the community has not widely adopted a standard format or the practice of openly
sharing data, some researchers and groups attempted to solve the first problem.
Most notably, Frenklach developed the PrIMe (\textbf{Pr}ocess \textbf{I}nformatics
\textbf{M}od\textbf{e}l) data format~\autocite{Frenklach:2007bm,You:2011hy} and an associated
online database, the PrIMe Data Warehouse~\autocite{PrIMe}. PrIMe files encode fundamental
combustion experimental data, along with kinetic models and calculated quantities,
using the Extensible Markup Language (XML) standard.

PrIMe represented an admirable first attempt at standardization, but a number of limitations
prevented wide adoption: the PrIMe standard does not require or support all information
needed to simulate an experiment, including a machine-readable definition of ignition or a
standard way to express the rapid compression machine (RCM) compression process; and
the XML format, intended more as a markup language than data format, presents a barrier
to creating and working with files.

More recently, Varga et al.~\autocite{Varga2015a,Varga2015b} developed the ReSpecTh standard,
an XML-based format that builds on PrIMe. While ReSpecTh adds important features, it still
suffers from the same usability issues as PrIMe.

We aim to develop a new, human- and machine-readable data standard for fundamental combustion
experiments, \ck{}, and offer tools for easily working with data encoded in this format.
Niemeyer recently introduced an initial version of the \ck{} format~\autocite{Niemeyer:2016wf},
and in this work we further formalize and develop the standard. We also discuss a related
Python software tool for validating and working with \ck{} files: \pk{}.
Our motivations are similar to some of those of the PrIMe and ReSpecTh teams, although
less ambitious in scope: enable easy sharing and use of data, for the (primary) purposes of
developing and validating predictive chemical kinetic models.
However, those efforts were generally less successful in being adopted---we hope to succeed
by developing more user-friendly data files and tools for interacting with them.
Furthermore, we plan to share all data and software openly, and cultivate a community of
user-developers.

>>>>

Interestingly, this contrasts the situation for chemical kinetic models, where the
CHEMKIN~\autocite{Kee:1996ck} format dominates. Competing standards such as the
Cantera~\autocite{Cantera:2.3.0} CTI format or FlameMaster~\autocite{FlameMaster:ref,FlameMaster:code}
lag behind considerably, although use of the former continually grows due to Cantera's
open availability.
\todo[inline, author=Bryan]{This reads more as a footnote, maybe consider putting it there instead?}
% Similarly, % with more space, discuss JANAF/NASA polynomial standard ?

Instead, such data is typically found in under-documented
comma-separated value files and Excel spreadsheets, or contained in PDF tables
rather than as supplementary material associated with a paper.
In the worst case, data is only available in figures and must be digitized by the use of software such as WebPlotDigitizer.
\todo{give a citation here}
% I think we're already stirring the pot enough, no need to call out
% specific people
% \todo{call out specific examples of these...?}{}
These practices limit the wide use of the valuable, extensive experimental data
available in the literature. Frenklach~\autocite{Frenklach:2007bm} further explained the
benefits of standardized, widely available data for combustion.
\todo[author=Bryan,inline]{It would maybe be good to give a bit of Frenklach's perspective here, rather than just throwing in the citation}

While the community has not widely adopted a standard format or the practice of openly \todo{this second bit feels like a slightly separate point, and it makes the second clause here a bit awkward}
sharing data, some researchers and groups attempted to solve the first problem.
Most notably, Frenklach developed the PrIMe (\textbf{Pr}ocess \textbf{I}nformatics
\textbf{M}od\textbf{e}l) data format~\autocite{Frenklach:2007bm,You:2011hy} and an associated
online database, the PrIMe Data Warehouse~\autocite{PrIMe}. PrIMe files encode fundamental
combustion experimental data, along with kinetic models and calculated quantities,
using the Extensible Markup Language (XML) standard.

PrIMe~\autocite{Frenklach:2007bm,You:2011hy} represented an admirable \todo{Maybe its just me, but this feels a bit sarcastic} first attempt at standardization, but a number of limitations
prevented wide adoption: the PrIMe standard does not require or support all information
needed to simulate an experiment, including a machine-readable definition of ignition or a
standard way to express the rapid compression machine (RCM) compression process; and
the XML format, intended more as a markup language than data format, presents a barrier
to creating and working with files.
\todo{Also because the standard was closed?}

More recently, Varga et al.~\autocite{Varga2015a,Varga2015b} developed the ReSpecTh standard,
an XML-based format that builds on PrIMe. While ReSpecTh adds important features, it still
suffers from the same usability issues as PrIMe.
\todo{slightly more detail?}

In this work, we present a new, open-source, human- and machine-readable data standard for fundamental combustion
experiments, \ck{}, and offer tools for easily working with data encoded in this format.
Niemeyer recently introduced an initial version of the \ck{} format~\autocite{Niemeyer:2016wf},
and in this work we further formalize and develop the standard. We also discuss a related
Python software tool for validating and working with \ck{} files: \pk{}.
Our motivations are similar those of the PrIMe and ReSpecTh teams, although
less ambitious in scope\todo{its not clear to me how this is less ambitious than the other projects}: the \ck{} project will enable easy sharing and use of fundamental combustion data, for the (primary) purposes of
developing and validating predictive chemical kinetic models.
Furthermore, we plan to share all data and software openly, and cultivate a community of
user-developers.

\todo[inline, author=Bryan]{We should probably also mention the CloudFlame project from KAUST.}
